# (PART) Step 3: Build  {-}

# Building app with `{golem}` {#stepbuild}

Now that the application is prototyped inside a `{golem}` skeleton, you can now work on the integration of the application. 

This parts encompasses the integration of the back-end/front-end, together, and working on the software engineering task of the application:

+ add and organize dependencies
+ creating and including sub-modules if necessary
+ organize utility functions and link them to the module they are used in
+ adding testing infrastructure 
+ linking to CI / CD services
+ check

### Add dependencies

Every time you add a new dependency, you can add it with:

```{r 11-step-by-step-build-1, eval = FALSE}
usethis::use_package("pkg")
```

Note that all the explicitly namespaced calls (_i.e_ `pkg::fun`) can be scraped using the `{attachment}` package:

```{r 11-step-by-step-build-2, eval = FALSE}
attachment::att_from_rscripts()
```

### Submodules and utility functions

Two functions can be called to add utility functions: 

```{r 11-step-by-step-build-3, eval = FALSE}
golem::add_fct( "helpers" ) 
golem::add_utils( "helpers" )
```

+ The first will create a `R/fct_helpers.R` file
+ The second will create a `R/utils_helpers.R` file

### Add tests

You can add more tests to your application with:

```{r 11-step-by-step-build-4, eval = FALSE}
usethis::use_test("app")
```


## Documentation and Code Coverage

### Vignette

Whenever you need to add a new Vignette, you can call:

```{r 11-step-by-step-build-5, eval = FALSE}
usethis::use_vignette("shinyexample")
```

The Vignettes can then be compiled with:

```{r 11-step-by-step-build-6, eval = FALSE}
devtools::build_vignettes()
```

### Code coverage

```{r 11-step-by-step-build-7, eval = FALSE}
usethis::use_travis()
usethis::use_appveyor()
usethis::use_coverage()
```

## Using `{golem}` dev functions

There is a series of tools to make your app behave differently whether it is in dev or prod mode. 
Notably, the `app_prod()` and `app_dev()` function look for the value of `options( "golem.app.prod" )`, or return TRUE if this option does not exist.

Setting this options at the beginning of your dev process allows to make your app behave in a specific way when you are in dev mode. 
For example, printing message to the console with `cat_dev()`.

```{r 11-step-by-step-build-8, eval = TRUE}
options( "golem.app.prod" = TRUE)
golem::cat_dev("hey\n")
options( "golem.app.prod" = FALSE)
golem::cat_dev("hey\n")
```

You can then make any function being "dev-dependant" with the `make_dev()` function: 

```{r 11-step-by-step-build-9, eval = TRUE}
log_dev <- golem::make_dev(log)
log_dev(10)
options( "golem.app.prod" = TRUE)
log_dev(10)
```
