# Common Application Caveats {#optim-caveat}

## Reactivity anti-patterns

### Reactivity is awesome... until it's not

Let's face it, reactivity is awesome... until it's not. 
Reactivity is a common source of confusion for beginners, and a common source of bugs and bottlenecks, even for seasoned Shiny developers. 
Most of the time, issues come from the fact that __there is too much reactivity__, i.e. we build apps where too much things happen, and some things are updated way more than they should, and computations are performed when they shouldn't. 

Of course, it's a nice feature to make everything react instantly to changes, but when building larger apps it's easy to create monsters, i.e complicated, messy reactive graphs where everything is updated too much and too often. 
Or worse, we generate endless reactive loops. 

And the more you work on your app, the more complex it gets, and the more you'll be likely to end up in the reactive inferno. 
In this section, we'll be speaking a little bit about reactivity and how to have more control on it, and about a way to share data across modules without relying on passing along reactive objects.

### `observe` vs `observeEvent`

One of the most common feature of reactive inferno is the use of `observe()` in cases where you should use `observeEvent`. 
Spoiler: you should try to use `observeEvent()` as much as possible, and avoid `observe()`as much as possible. 
At first, `observe()` seems easier to implement, and feels like a shortcut as you don't have to think about what to react to: everything gets updated without you thinking about it.
But the truth is that this stairway doesn't lead to heaven. 

Let's stop and think about `observe()` for a minute. 
This function updates __every time a reactive object it contains is invalidated__. 
Yes, this works well if you have a small amount of reactive objects in the observer, but that gets tricky whenever you start adding things inside your `observe()`, as you might be launching a computation 10 times if your reactive scope contains 10 reactive objects that are somehow invalidated in chain. 
And believe me, we've seen pieces of code where the `observe()` contains hundreds of lines of code, with reactives objects all over the place.

For example, let's start with that:

```{r common-app-caveats-1, eval = FALSE}
## DON'T DO GLOBAL VARIABLES, IT'S JUST TO SIMPLIFY THE EXAMPLE
i <- 0
library(shiny)
library(cli)
ui <- function(request){
  tagList(
    textInput("txt", "txt")
  )
}

server <- function(input, output, session){
  observe({
    i <<- i + 1
    cat_rule(as.character(i))
    print(input$txt)
  })
}

shinyApp(ui, server)
```

Oh, and then, let's add a small selectInput

```{r common-app-caveats-2, eval = FALSE}
i <- 0
library(shiny)
library(cli)
ui <- function(request){
  tagList(
    textInput("txt", "txt"), 
    selectInput("tolower", "casse", c("lower", "upper"))
  )
}

server <- function(input, output, session){
  observe({
    i <<- i + 1
    cat_rule(as.character(i))
    if (input$tolower == "lower") {
      print(tolower(input$txt))
    } else  {
      print(tolower(input$txt))
    }
  })
}

shinyApp(ui, server)
```

And, as time goes by, we add another control flow to our `observe()`: 

```{r common-app-caveats-3, eval = FALSE}
i <- 0
library(shiny)
library(cli)
library(stringi)
ui <- function(request){
  tagList(
    textInput("txt", "txt"), 
    selectInput("tolower", "casse", c("lower", "upper")), 
    checkboxInput("rev", "reverse")
  )
}

server <- function(input, output, session){
  observe({
    i <<- i + 1
    cat_rule(as.character(i))
    if (input$rev){
      x <- stri_reverse(input$txt)
    } else {
      x <- input$txt
    }
    if (input$tolower == "lower"){
      print(tolower(x))
    } else {
      print(tolower(x))
    }
  })
}

shinyApp(ui, server)
```

And it would be nice to keep the selected values into a reactive list, so that we can reuse it elsewhere. 
And maybe you would like to add a checkbox so that the logs are printed to the console only if checked.

```{r common-app-caveats-4, eval = FALSE}
i <- 0
library(shiny)
library(cli)
library(stringi)
ui <- function(request){
  tagList(
    textInput("txt", "txt"), 
    selectInput("tolower", "casse", c("lower", "upper")), 
    checkboxInput("rev", "reverse")
  )
}

server <- function(input, output, session){
  r <- reactiveValues()
  observe({
    i <<- i + 1
    cat_rule(as.character(i))
    if (input$rev) {
      r$x <- stri_reverse(input$txt) 
    } else {
      r$x <- input$txt
    }
    if (input$tolower == "lower"){
      r$x <- tolower(r$x)
    } else {
      r$x <- toupper(r$x)
    }
  })
}

shinyApp(ui, server)
```

Ok, now can you tell how many potential invalidation points we've got here? 
Three: whenever `input$txt`, `input$rev` or `input$tolower` change. 
Of course, three is not that much, but you get the idea.

Let's pause a minute and think about why we use `observe()` here. 
To update the values inside `r$x`, yes. 
But do we need to use `observe()` for, say, updating `r$x` under dozens of conditions, each time the user types a letter? 
I don't think so. 

We generally want our observer to update its content under a small, controlled number of inputs, i.e. with a controlled number of invalidation points. 
And, what we often forget is that users don't type/select correctly on the first try. 
No, they usually try and miss, restart, change things, amplifying the reactivity "over-happening". 

Moreover, long `observe()` statements are hard to debug, and they make collaboration harder when the trigger to the observe logic can potentially lives anywhere between line one and line 257 of your `observe()`. That's why (well, in 99% of cases), it's safer to go with `observeEvent`, as it allows to see at a glanche what are the condition under which the content is invalidated and re-evalutaed. 
Then, if a reactive context is invalidated, __you know why__.

### Building triggers and watchers

// TODO 

So the general pattern is that one: we create "flags" with an `init` function, trigger these "flags" with `trigger()`, and wherever we want these flags to invalidate the reactive context, we `watch()` these flags. 

The idea here is to get a full control over the reactive flow: we only invalidate contexts when we want, making the general flow of the app more predictable.

```{r common-app-caveats-5}
init <- function(name, session = getDefaultReactiveDomain()){
  session$userData[[name]] <- reactiveVal(0)
}

trigger <- function(name, session = getDefaultReactiveDomain()){
  session$userData[[name]](
    session$userData[[name]]() + 1
  )
}

watch <- function(name, session = getDefaultReactiveDomain()){
  session$userData[[name]]()
}
```

And, bonus, as these functions use the `session` object, they are available across all modules. That also means that you can easily trigger an event inside a model from another one. 

### Using R6 as a data storage

One pattern we've also been playing with is storing the app business logic inside of one or more R6 objects. 
Why would we want to do that?

#### Sharing data accross module

Sharing an R6 object makes it simpler but without the complexity of sharing & handling reactive objects. 

So basically, the idea is to hold the whole logic of your data reading / cleaning / processing / outputing inside an R6 class. 
You'll then initiate an instance of this class, then pass this object through your modules and submodules. 

```{r common-app-caveats-6, eval = FALSE}
nameui <- function(id){
  ns <- NS(id)
  tagList(
    # [...]
  )
}

name <- function(input, output, session, obj){
  ns <- session$ns
  output$that <- renderThis({
    obj$compute()
  })
}

library(shiny)
ui <- function(request){
  tagList(
    nameui("nameui")
  )
}

server <- function(
  input, 
  output, 
  session
){
  obj <- MyDataProcess$new()
  callModule(name, "nameui", obj)
  
}

shinyApp(ui, server)

```

Of course, this makes sense if you're combining it with the trigger / watch from before :) 

#### Get sure it's tested

During the process of building a robust Shiny app, we strongly suggest that you test as many things as you can. 
This is where using an R6 for your business logic of your app makes sense: this allows you to build the whole testing of your application logic outside of any reactive context: you simply build unit tests just as any other function. 

## R does too much 

### Rendering UI from server side 

There are many reasons we would want to change things on the UI based on what happens in the server: changing the choices of a `selectInput()` based on the columns of a table which is uploaded by the user, showing and hidding pieces of the app according to an environment variable, allow the user to create an undeterminate amount of inputs, etc. 

Chances are that to do that, you've been using the `uiOutput()` & `renderUI()` functions from `{shiny}`.
Even if convenient, and the functions of choice in some specific context, this couple makes R do a little bit too much: you're making R regenerate the whole UI component instead of changing only what you need. 
Plus, you'll create a code that is harder to reason about, as we are used to have the UI parts in the UI functions (but that's not related to performance).

Here are three strategies to code without `uiOutput()` & `renderUI()`.

#### Implement UI events in JavaScript

> Mixing languages is better than writing everything in one, if and only if using only that one is likely to overcomplicate the program.

`r right_link("Applying the Unix Philosophy - The Art of Unix Programming", "http://www.catb.org/~esr/writings/taoup/html/ch01s08.html")`

We'll see in the last chapter of this book how you can integrate JS inside your Shiny app, and how event basic functions can be useful in making your app server lighter. 
For example, compare: 

```{r}
library(shiny)
ui <- function(){
  tagList(
    actionButton("change", "show/hide graph", onclick = "$('#plot').toggle()"), 
    plotOutput("plot")
  )
}

server <- function(
  input, 
  output, 
  session
){
  output$plot <- renderPlot({
    cli::cat_rule("Rendering plot")
    plot(iris)
  })
}

shinyApp(ui, server)
```

to

```{r}
library(shiny)
ui <- function(){
  tagList(
    actionButton("change", "show/hide graph"), 
    plotOutput("plot")
  )
}

server <- function(
  input, 
  output, 
  session
){
  
  r <- reactiveValues(plot = iris)
  
  observeEvent( input$change , {
    if (input$change %% 2 == 0){
      r$plot <- iris
    } else {
      r$plot <- NULL
    }
  })
  
  output$plot <- renderPlot({
    cli::cat_rule("Rendering plot")
    req(r$plot)
    plot(r$plot)
  })
  
}

shinyApp(ui, server)
```

The result is the same, but the first version is shorter and easier to reason about: we have one bouton, and the behavior of the button is contained into itself. 
And, on top of being harder to maintain, the second solution redraws the plot everytime the `reactiveValues` is updated, making R compute way more than it should. 

#### `update*` inputs

Almost every Shiny inputs, even the custom ones from packages, come with an `update` function that allows to change the input content from the server side, instead of recreating entirely. 
For example, here is a way to update the content of a `selectInput` from the server side:

```{r}
library(shiny)
ui <- function(){
  tagList(
    selectInput("species", "Species", choices = NULL), 
    actionButton("update", "Update")
  )
}

server <- function(
  input, 
  output, 
  session
){
  observeEvent( input$update , {
    spc <- unique(iris$Species)
    updateSelectInput(
      session, 
      "species", 
      choices = spc, 
      selected = spc[1]
    )
  })
  
}

shinyApp(ui, server)
```

Here, this makes the code easier to reason about as the `selectInput` is where it should be: inside the UI. 
Plus, with the `update` method, we are only changing what is needed, not re-generating the whole input. 
 
#### `insertUI` and `removeUI` 

Another way to dynamically change what is in the UI is with `insertUI()` and `removeUI()`. 
It's more global than the solution we've seen before with setting the `reactiveValue` to `NULL` or to a value, as it allows to target a larger UI element: we can insert or remove the whole input, instead of having the DOM element present but empty. 
This technic allows to have a lighter DOM: div which are not rendered are not generated empty, they are simply not there. 

Two things to note concerning this method, though: 

+ Removing an element from the app will not delete the input from the input list. 
In other word, if you have `selectInput("x", "x")`, and that you remove this input using `removeUI()`, you will still have `input$x` in the server. 

+ Both these functions take a `jQuery` selector to select the element in the UI. 
We will introduce these selectors in the last chapter of this book.

### Too Much Data in Memory

If you are building a Shiny application, there is a great chance you are building it to analyse data. 
If you are dealing with large datasets, you should consider deporting the data handling and computation to an external database system: for example to an SQL database. 
Why? Because these system has been created to handle and manipulate data on disk: in other words it will allow you to perform operation on your data without having to clutter R memory with large dataset. 
For example, if you have a `selectInput()` that is used to perform a filter on a dataset, you can do that straight inside SQL. 
Well, almost any operation you can do today in `{dplyr}` would be doable with an SQL backend, and that's the purpose of the `{dbplyr}` package: translate `{dplyr}` code into SQL.

If using a database as a backend seems a little bit far-fetched to you, that's how it's done in most programming languages: if you are building a web app with NodeJS or Python and need to interact with data, nothing will be stored in RAM: you will be relying on an external database to store your data.  

## Reading data

// TODO
