# Common Application Caveats {#optim-caveat}

## Reactivity anti-patterns

### Reactivity is awesome... until it's not

Let's face it, reactivity is awesome... until it's not. 
Reactivity is a common source of confusion for beginners, and a common source of bugs and bottlenecks, even for seasoned Shiny developers. 
Most of the time, issues come from the fact that __there is too much reactivity__, i.e. we build apps where too much things happen, and some things are updated way more than they should, and computations are performed when they shouldn't. 

Of course, it's a nice feature to make everything react instantly to changes, but when building larger apps it's easy to create monsters, i.e complicated, messy reactive graphs where everything is updated too much and too often. 
Or worse, we generate endless reactive loops. 

And the more you work on your app, the more complex it gets, and the more you'll be likely to end up in the reactive inferno. 
In this section, we'll be speaking a little bit about reactivity and how to have more control on it, and about a way to share data across modules without relying on passing along reactive objects.

### `observe` vs `observeEvent`

One of the most common feature of reactive inferno is the use of `observe()` in cases where you should use `observeEvent`. 
Spoiler: you should try to use `observeEvent()` as much as possible, and avoid `observe()`as much as possible. 
At first, `observe()` seems easier to implement, and feels like a shortcut as you don't have to think about what to react to: everything gets updated without you thinking about it.
But the truth is that this stairway doesn't lead to heaven. 

Let's stop and think about `observe()` for a minute. 
This function updates __every time a reactive object it contains is invalidated__. Y
es, this works well if you have a small amount of reactive objects in the observer, but that gets tricky whenever you start adding things inside your `observe()`, as you might be launching a computation 10 times if your reactive scope contains 10 reactive objects that are somehow invalidated in chain. 
And believe me, we've seen pieces of code where the `observe()` contains hundreds of lines of code, with reactives objects all over the place.

For example, let's start with that:

```{r common-app-caveats-1, eval = FALSE}
## DON'T DO GLOBAL VARIABLES, IT'S JUST TO SIMPLIFY THE EXAMPLE
i <- 0
library(shiny)
library(cli)
ui <- function(request){
  tagList(
    textInput("txt", "txt")
  )
}

server <- function(input, output, session){
  observe({
    i <<- i + 1
    cat_rule(as.character(i))
    print(input$txt)
  })
}

shinyApp(ui, server)
```

Oh, and then, let's add a small selectInput

```{r common-app-caveats-2, eval = FALSE}
i <- 0
library(shiny)
library(cli)
ui <- function(request){
  tagList(
    textInput("txt", "txt"), 
    selectInput("tolower", "casse", c("lower", "upper"))
  )
}

server <- function(input, output, session){
  observe({
    i <<- i + 1
    cat_rule(as.character(i))
    if (input$tolower == "lower") {
      print(tolower(input$txt))
    } else  {
      print(tolower(input$txt))
    }
  })
}

shinyApp(ui, server)
```

And, as time goes by, we add another control flow to our `observe()`: 

```{r common-app-caveats-3, eval = FALSE}
i <- 0
library(shiny)
library(cli)
library(stringi)
ui <- function(request){
  tagList(
    textInput("txt", "txt"), 
    selectInput("tolower", "casse", c("lower", "upper")), 
    checkboxInput("rev", "reverse")
  )
}

server <- function(input, output, session){
  observe({
    i <<- i + 1
    cat_rule(as.character(i))
    if (input$rev){
      x <- stri_reverse(input$txt)
    } else {
      x <- input$txt
    }
    if (input$tolower == "lower"){
      print(tolower(x))
    } else {
      print(tolower(x))
    }
  })
}

shinyApp(ui, server)
```

And it would be nice to keep the selected values into a reactive list, so that we can reuse it elsewhere. 
And maybe you would like to add a checkbox so that the logs are printed to the console only if checked.

```{r common-app-caveats-4, eval = FALSE}
i <- 0
library(shiny)
library(cli)
library(stringi)
ui <- function(request){
  tagList(
    textInput("txt", "txt"), 
    selectInput("tolower", "casse", c("lower", "upper")), 
    checkboxInput("rev", "reverse")
  )
}

server <- function(input, output, session){
  r <- reactiveValues()
  observe({
    i <<- i + 1
    cat_rule(as.character(i))
    if (input$rev) {
      r$x <- stri_reverse(input$txt) 
    } else {
      r$x <- input$txt
    }
    if (input$tolower == "lower"){
      r$x <- tolower(r$x)
    } else {
      r$x <- toupper(r$x)
    }
  })
}

shinyApp(ui, server)
```

Ok, now can you tell how many potential invalidation points we've got here? 
Three: whenever `input$txt`, `input$rev` or `input$tolower` change. 
Of course, three is not that much, but you get the idea.

Let's pause a minute and think about why we use `observe()` here. 
To update the values inside `r$x`, yes. 
But do we need to use `observe()` for, say, updating `r$x` under dozens of conditions, each time the user types a letter? 
I don't think so. 

We generally want our observer to update its content under a small, controlled number of inputs, i.e. with a controlled number of invalidation points. 
And, what we often forget is that users don't type/select correctly on the first try. 
No, they usually try and miss, restart, change things, amplifying the reactivity "over-happening". 

Moreover, long `observe()` statements are hard to debug, and they make collaboration harder when the trigger to the observe logic can potentially lives anywhere between line one and line 257 of your `observe()`. That's why (well, in 99% of cases), it's safer to go with `observeEvent`, as it allows to see at a glanche what are the condition under which the content is invalidated and re-evalutaed. 
Then, if a reactive context is invalidated, __you know why__.

### Building triggers and watchers

// TODO 

So the general pattern is that one: we create "flags" with an `init` function, trigger these "flags" with `trigger()`, and wherever we want these flags to invalidate the reactive context, we `watch()` these flags. 

The idea here is to get a full control over the reactive flow: we only invalidate contexts when we want, making the general flow of the app more predictable.

```{r common-app-caveats-5}
init <- function(name, session = getDefaultReactiveDomain()){
  session$userData[[name]] <- reactiveVal(0)
}

trigger <- function(name, session = getDefaultReactiveDomain()){
  session$userData[[name]](
    session$userData[[name]]() + 1
  )
}

watch <- function(name, session = getDefaultReactiveDomain()){
  session$userData[[name]]()
}
```

And, bonus, as these functions use the `session` object, they are available across all modules. That also means that you can easily trigger an event inside a model from another one. 

### Using R6 as a data storage

One pattern we've also been playing with is storing the app business logic inside of one or more R6 objects. 
Why would we want to do that?

#### Sharing data accross module

Sharing an R6 object makes it simpler but without the complexity of sharing & handling reactive objects. 

So basically, the idea is to hold the whole logic of your data reading / cleaning / processing / outputing inside an R6 class. 
You'll then initiate an instance of this class, then pass this object through your modules and submodules. 

```{r common-app-caveats-6, eval = FALSE}
nameui <- function(id){
  ns <- NS(id)
  tagList(
    # [...]
  )
}

name <- function(input, output, session, obj){
  ns <- session$ns
  output$that <- renderThis({
    obj$compute()
  })
}

library(shiny)
ui <- function(request){
  tagList(
    nameui("nameui")
  )
}

server <- function(
  input, 
  output, 
  session
){
  obj <- MyDataProcess$new()
  callModule(name, "nameui", obj)
  
}

shinyApp(ui, server)

```

Of course, this makes sense if you're combining it with the trigger / watch from before :) 

#### Get sure it's tested

During the process of building a robust Shiny app, we strongly suggest that you test as many things as you can. 
This is where using an R6 for your business logic of your app makes sense: this allows you to build the whole testing of your application logic outside of any reactive context: you simply build unit tests just as any other function. 

## R does too much 

// TODO 

## Reading data

// TODO
