# A Gentle Introduction to CSS {#css}

## What is CSS

### About CSS

CSS, for `Cascading Style Sheets`, is one of the most commonly used technology that powers the web today, along wîth HTML and JavaSCript. 
HTML is a series of tags that define your web page structure, and JavaScript is a language that allows you to manipulate the page (well, it can do a lot more than that, but we're simplifying to make it understandable). 
CSS is what handles the design, i.e. the visual interface of the webpage: the color of the header, the font, the background, and things like that (again, we're simplifying for the sake of clarity). 
Every HTML element is shown with a default style: for example all `<h1>` have the size `2em` and are in bold, and `<strong>` is in bold.
This is what we'll be using CSS for: overriding these default values.

In Shiny, there is a default CSS: the one from Bootstrap. 
As you can see if you have created a `fluidPage()` before, there is already styling applied. 

Compare: 

(no `fluidPage`)

```{r css-1}
library(shiny)
ui <- function(request){
  tagList(
    h1("Hey"), 
    h2("You"), 
    p("You rock!"), 
    selectInput("what", "Do you", unique(iris$Species))
  )
}

server <- function(
  input, 
  output, 
  session
){

}

shinyApp(ui, server)
```

to (with `fluidPage`):

```{r css-2}
library(shiny)
ui <- function(request){
  fluidPage(
    h1("Hey"), 
    h2("You"), 
    p("You rock!"), 
    selectInput("what", "Do you", unique(iris$Species))
  )
}

server <- function(
  input, 
  output, 
  session
){

}

shinyApp(ui, server)
```

Yes, that's subtle, but you can see how it makes the difference on larger apps.

## Getting started with CSS 

// TODO

## Integrate CSS files to your Shiny App

// TODO
