# Introduction to `{golem}` {#golem}

Ok, that’s a lot of things to process. 
Is there a tool that can help us simplify this workflow? 
Of course there is, and it’s called `{golem}`.

It can be found at https://github.com/ThinkR-open/golem

The stable release can be found on CRAN, and installed with:

```{r eval = FALSE}
install.packages("golem")
```

A more up to date (but maybe less stable) version of `{golem}` can also be found on GitHub that you can install with:

```{r eval = FALSE}
remotes::install_github("Thinkr-open/golem")
```

`{golem}` is an R package that implements an opinionated framework for building production-ready Shiny apps. 
It all starts with an RStudio project, which contains a predefined setup for building your app. 
The idea is that with `{golem}`, you don’t have to focus on the foundation of your app, and can spend your time thinking about what you want to do, not about how to do it. 
It’s built on top of the working process we’ve developed at ThinkR, and tries to gather in one place the functions and tools we’ve created for building applications designed for production.

## Getting started with `{golem}`

Note before using `{golem}`: 

- A `{golem}` app is contained inside a package, so knowing how to build a package is highly recommended. 

- A `{golem}` app works better if you are working with `shiny modules`, so knowing how modules work is heavily recommended.

In the rest of the book, we'll assume you're working in RStudio. 

## Understanding `{golem}` package structure

// TODO
