# Using JavaSript {#optimjs}

```{r include = FALSE}
library(magrittr)
golem::add_js_handler("plop", "golemexample", open = FALSE)
```


In its core, building a Shiny app is building a JavaScript app that can talk with an R session. 
This process is hidden from the common Shiny developer, who usually thinks that everything is done in R. 
In a sense, it's true, but when you are writting UI elements in Shiny what you are doing is building a series of HTML tags with JavaScript events, that will communicate with R, read results, and update the page the app-user sees. 
What happens under the hood is a little bit complex and  than that but the general idea is there: R talks to your browser through a websocket, and this very same browser talks to R through the same websocket.

// TODO: create here a simple Flowchart 
// R -> (WebSocket) -> JS 
// R <- (WebSocket) <- JS 

It's important to note here that the communication happens in both ways: from R to JavaScript, and from JavaScript to R. 
In fact, what we do when we write a piece of code like `sliderInput("plop", "this", 1, 10, 5)`, what we are doing is creating a binding between JavaScript and R, where JavaScript listens to any event happening on the slider with the id `"plop"`, and whenever JavaScript detects that something happens to this element, its value is sent back to R. 
With `output$bla <- renderPlot({})`, what we are doing is making the two communicates the other way around: we are telling JavaScript to listen to any incoming data from R for the `id` `"bla"`, and whenever JavaScript sees incoming data from R, it puts it into the proper HTML element. 

So even if everything happens in R, we __are__ writting a web app, from R. 
Once you've realized that, the possibilities are endless: in fact almost anything doable in a web app UI can be done in Shiny. 
What this also implies is that getting (even a little bit) better at writting HTML, CSS and especially JavaScript will make your app better, lighter, and more performant. 
Indeed, JavaScript for example is a language which has been designed to interact with a web page: change element appearances, hide and show things, cick somewhere, send alerts... 
What that means is that knowing just enought JavaScript can improve the quality of your app: especially when you've been using R to build more or less complex interface rendering: conditional pannels, simulating a button click, hide and showing elements are for example good examples of where you should be using JavaScript instead of building conditionnal `renderUI` or `insertUI` in your server.

This is what this part aims to do: giving you just enough JavaScript knowledge to lighten your Shiny App, and improve the global user and developer experience.
In this part, we'll see how to make R and JavaScript communicate with each other. 
We'll also see how to do some simple "client-side only" optimizations, and explore common patterns for JavaScript in Shiny. 
Finally, we'll quickly present some of the functions from `{golem}` which are built on top of JavaScript. 

> Note that this chapter is not supposed to be a comprehensive JavaScript course. 
External resources are linked at the botton of this chapter if you want to dive deeper into JavaScript.

## JavaScript <-> Shiny communication 

### From R to JavaScript 

Inside this skeleton:

```{r echo = FALSE}
readLines("golemexample/inst/app/www/plop.js") %>%
  glue::as_glue()
```

> `golem::add_js_handler()`

Called from server-side with:


```{r eval = FALSE}
session$sendCustomMessage("fun", list())
```

`list()` -> JSON. 

Example:

- JavaScript 

```{js, eval = FALSE, echo = TRUE}
Shiny.addCustomMessageHandler('monalerte', function(arg) {
  alert(arg.message);
  alert(arg.message2);
})
```

- R 

```{r eval = FALSE}
session$sendCustomMessage(
  "monalerte", 
  list(
    message = "Mon premier message", 
    message2 = "Mon deuxième message"
  )
)
```



### From JavaScript to R

```{js, eval = FALSE, echo = TRUE}
Shiny.setInputValue("rand", Math.random())
```

```{r eval = FALSE}
observeEvent( input$rand , {
  print( input$rand )
})
```

## Client-side Optimization

> `golem::add_js_file()`

### About jQuery & jQuery selectors

The `jQuery` framework is natively included in Shiny

> jQuery is a fast, small, and feature-rich JavaScript library. It makes things like HTML document traversal and manipulation, event handling, animation, and Ajax much simpler with an easy-to-use API that works across a multitude of browsers. With a combination of versatility and extensibility, jQuery has changed the way that millions of people write JavaScript.

It comes with the notion of selectors:

+ `#plop`: the element with the id `plop`

+ `.pouet`: elements of class `pouet`

+ `"button:contains('this')"`: buttons with a text containing `'this'`

Note that in html, tags contains attributes. For example:

``` html
<a href = "https://thinkr.fr" data-value = "panel2">ThinkR</a>
```

contains `href` & `data-value`. 
You can refer to these attributes with `[]` after the tag name.

+ `a[href = "https://thinkr.fr"]`: link with `href` being `https://thinkr.fr`

+ `a[data-value="panel2"]`: link with  `data-value` being `"panel2"`

#### div, class, id

+ div == "bloc" in an HTML page

+ Can have an `id`,which has to be unique, references with `#nomid`

+ It can have a class, class can be added to several elements. Referenced with `.nomclass`

### Adding attributes to HTML elements

+`onclick`

```{r eval = FALSE}
actionButton(
  "plop", "plop", 
  onclick = "alert('hello world')"
)
```

+ `shiny::tagAppendAttributes`

```{r eval = FALSE}
plotOutput(
  "plop"
  ) %>% tagAppendAttributes(
    onclick = "alert('hello world')"
  )
```


```{r eval = FALSE}
library(shiny)
ui <- function(request){
  fluidPage(
    plotOutput(
      "plop"
    ) %>% tagAppendAttributes(
      onclick = "alert('hello world')"
    )
  )
}

server <- function(input, output, session){
  output$plop <- renderPlot({
    plot(iris)
  })
}

shinyApp(ui, server)
```

## Common JavaScript patterns for Shiny

+ `$('#id').show()` 

+ `$('#id').hide()` 

+ `alert("message")` 

+ `var x = prompt("this", "that")` 

+ `$('#id').css('color', 'blue');` 

+ `$("#id").text( "this" );` 

+ `$("#id").remove();` 


## About `{golem}` js functions

`{golem}` comes with a series of JavaScript functions that you can call from the server. 
These functions are added by default with `golem::activate_js()` in `app_ui`.


Then they are called with `golem::invoke_js("function", "ui_element")`.

This `ui_element` define the UI element to interact with. 
It can be a full jQuery selector, an id or a class.

### `golem::activate_js()`

+ `showid` & `hideid`, `showclass` & `hideclass` show and hide elements through their id or class.

```{r eval = FALSE}
session$sendCustomMessage("showid", ns("plot"))
```

+ `showhref` & `hidehref` hide and show a link by trying to match the `href` content.

```{r eval = FALSE}
session$sendCustomMessage("showhref", "panel2")
```

+ `clickon` click on the element. You have to use the full jQuery selector. 

+ `show` & `hide` show and hide elements, using the full jQuery selector.

## Learn more about JavaScript

