# Using JavaScript {#optimjs}

```{r include = FALSE}
knitr::opts_chunk$set( comment = "", eval = FALSE)
```

In its core, __building a Shiny app is building a JavaScript app__ that can talk with an R session. 
This process is invisible for most Shiny developers, who usually do everything with R. 
And in the end, it's the case: most of the Shiny apps out there are written with R. 

But in fact, when you are writing UI elements in Shiny, __what you are actually doing is building a series of HTML tags, which are then linked to JavaScript events__. 
Later on, when the app is running, these JavaScript events will communicate with R, in the sense that they will send data to R, and receive data from R. 
Most of the time, when these JavaScript events are receiving data, they modify the page the user sees.
What happens under the hood is a little bit complex and out of scope for this book, but the general idea is there: R talks to your browser through a web socket (that you can imagine as a small "phone line" with both softwares listening on one hand^[
See this [dev.to](https://dev.to/buzzingbuzzer/comment/g0g) post for a quick introduction to the general concept of web socket
]), and this browser talks to R through the same web socket.

// TODO: create here a simple Flowchart 

// R -> (Web Socket) -> JS 

// R <- (Web Socket) <- JS 

It's important to note here that the __communication happens in both ways__: from R to JavaScript, and from JavaScript to R. 
In fact, when we write a piece of code like `sliderInput("plop", "this", 1, 10, 5)`, what we are doing is creating a binding between JavaScript and R, where JavaScript listens to any event happening in the browser on the slider with the id `"plop"`, and whenever JavaScript detects that something happens to this element, something (most of the time its value) is sent back to R. 
With `output$bla <- renderPlot({})`, what we are doing is making the two communicates the other way around: we are telling JavaScript to listen to any incoming data from R for the `id` `"bla"`, and whenever JavaScript sees incoming data from R, it puts it into the proper HTML tag (here, JavaScript inserts in the page the image received from R). 

So even if everything is written with R, we __are__ writing a web application, i.e. HTML, CSS and JavaScript elements. 
Once you've realized that, the possibilities are endless: in fact almost anything doable in a "classic" web app can be done in Shiny. 
What this also implies is that getting (even a little bit) better at writing HTML, CSS and especially JavaScript will make your app better, lighter, and more user friendly, as JavaScript is a language that has been designed to interact with a web page: change element appearances, hide and show things, click somewhere, show alerts and prompts... 
__Knowing just enough JavaScript can improve the quality of your app__: especially when you've been using R to build some complex UI renderings: think conditional panels, simulating a button click from the server, hide and show elements... 
All these things are good examples of where you should be using JavaScript instead of building more or less complex `renderUI` or `insertUI` patterns in your server.
A good rule to live

Moreover, the number of JavaScript libraries available on the web is tremendous ; and the good news is that Shiny has everything it needs to bundle external JavaScript libraries inside your application^[
This can also be done by wrapping a JS libraries inside a package, which will later be used inside an application. 
See for example [`{glouton}`](https://github.com/ColinFay/glouton), which is a wrapper around the [`js-cookie`](https://github.com/js-cookie/js-cookie) JavaScript library.
].

This is what this part of the book aims at: giving you just enough JavaScript knowledge to lighten your Shiny App, in order to improve the global user and developer experience.
In this part, we'll first see some JavaScript basics, which can be used "client-side" only, i.e. only in your browser, or by making R & JS communicate with each other. 
In this part, we'll also explore common patterns for JavaScript in Shiny. 
Finally, we'll quickly present some of the functions from `{golem}` which are built on top of JavaScript. 

> Note that this chapter is not supposed to be a comprehensive JavaScript course. 
External resources are linked all throughout this chapter and at the end if you want to dive deeper into JavaScript.

## A quick introduction to JavaScript

JavaScript is a programming language which has been designed to work in the browser^[
You can now work with JavaScript in a server with Node.JS, but this is out of scope of this book. 
See linked resources to learn more.
].
There are three ways to include the JavaScript code inside your web app: 

+ As an external file, which is served to the browser alongside your main application page
+ Inside a `<script>` HTML tag inside your page
+ On a specific tag, for example by adding an `onclick` event straight on the tag

> Note that the good practice when it comes to include JavaScript is to add the code inside an external file.

If you're working with `{golem}`, including a JavaScript file is performed using two functions: 

+ `golem::add_js_file("name")`, which adds a standard JavaScript file, i.e. one which is not meant to be used to communicate with R.
We'll see in the first part of this chapter how to add JavaScript code there.
+ `golem::add_js_handler("name")`, which creates a file with a skeleton for Shiny handlers. 
We'll see this second type of elements in the `JavaScript <-> Shiny communication` part.

OK, good, but what do we do now?

### Understanding html, class, and id

You have to think of a web page as a tree, where the top of the webpage is the root node, and every element in the page is a node in this tree (this tree is called a DOM, for Document Object Model). 
__You can work on any of these HTML node with JavaScript__: modify it, bind and/or listen to events, hide and show it...
But first, __you have to find a way to identify these elements__: either as a group of elements or as a unique element inside the whole tree. 
That's what html semantic elements, classes and ids are made for.
Consider this piece of code:

```{r echo = TRUE, eval = FALSE}
library(shiny)
fluidPage(
  titlePanel("Hello Shiny"), 
  actionButton("go", "go")
)
```

```
<div class="container-fluid">
  <h2>Hello Shiny</h2>
  <button id="go" type="button" class="btn btn-default action-button">go</button>
</div>
```


This `{shiny}` code creates a piece of HTML code, which contains three nodes: a `div` with a specific class (a BootStrap container), an `h2`, which is a level two header, and a button which has an id and a class.
Both are included into the `div`. 
Let's detail what we've got here:

+ HTML tags, which are the building blocks of the "tree": here, `div`, `h2` and `button` are HTML tags.
+ The button has an `id`, which is short for "identifier". 
This id has to to be unique: it's this reference that allows to spot this exact element, and more specifically it's this id that allows JavaScript and R to talk to each other: if you click on a button, you have to be sure that it's this specific button, and only that one.
+ Elements can have a class, which can apply to multiple elements. 
This can be used in JavaScript, but it's also very useful for styling elements in CSS. 

### About jQuery & jQuery selectors

The `jQuery` framework is natively included in Shiny.

> jQuery is a fast, small, and feature-rich JavaScript library. It makes things like HTML document traversal and manipulation, event handling, animation, and Ajax much simpler with an easy-to-use API that works across a multitude of browsers. 

`r right_link("jQuery home page", "https://jquery.com/")`

jQuery is a very popular JavaScript library which is designed to manipulate the DOM, its events and its elements. 
It can be used to do a lot of things, like hide and show, change class, click somewhere... 
And to be able to do that, it comes with the notion of selectors, which will be put between `$()`.
You can use, for example:

+ `$("#plop")` to refer to the element with the id `plop` 

+ `$(".pouet")` to refer to element(s) of class `pouet`

+ `$("button:contains('this')")` to refer to the buttons with a text containing `'this'`

You can also use special HTML attributes, which are specific to a tag.
For example, the following HTML code:

``` html
<a href = "https://thinkr.fr" data-value = "panel2">ThinkR</a>
```

contains the `href` & `data-value` attributes.
You can refer to these with `[]` after the tag name.

+ `$("a[href = 'https://thinkr.fr'])` refers to link(s) with `href` being `https://thinkr.fr`

+ `$('a[data-value="panel2"]')` refers to link(s) with  `data-value` being `"panel2"`

All these selectors, along many others, are __used to identify one or more node(s) in the big tree which is a web page__. 
Once we've identified these elements, we can either extract or change data contained in these nodes, or launch methods contained inside these nodes.
Indeed JavaScript, as R, can be used as a functional language, but most of what we do is done in an object-oriented way. 
In other words, you'll interact with objects from the web page, and these objects contain data and methods. 

Note that this is not specific to jQuery: elements can also be selected with standard JavaScript. 
jQuery has the advantage of simplifying selections and actions, and to be cross-platform, making it easier to ship applications that can work on all major browsers. 
And, it comes with Shiny for free! 

## Client-side JavaScript

It's hard to give an exhaustive list of what you can do with JavaScript inside Shiny. 
As a Shiny app is part JavaScript, part R, once you've got a good grasp of JavaScript you can quickly enhance any of your applications.
That being said, there are some common things that can be done that would allow you to immediately optimize your application: i.e. small JavaScript functions that will prevent you from writing complex algorithmic logic in your application server. 

### Common patterns

+ `$('#id').show();` and `$('#id').hide();` show and hide one or more elements that match the given selector.
For example, this can be use to replace: 

```{r eval = FALSE}
output$ui <- renderUI({
  if (this){
    tags(...)
  } else {
    NULL
  }
})
```

+ `alert("message")` uses the built-in alert-box mechanism from the user's browser (i.e., the `alert()` function is not part of jQuery but it's built inside the user's browser).
It works well as it relies on the browser instead of relying on R or on specific JavaScript library. 
You can use this functionality to replace a call to `{shinyalert}`: the result is a little bit less aesthetically pleasing, but that's easier to implement and maintain.

+ `var x = prompt("this", "that");` this function open the built-in prompt, which is a text area where the user can input text. 
With this code, when the user clicks on "OK", the text is stocked in a variable `x`, which you can then send back to R (see further part down this chapter for more info about how to do that). 
This can replace something like : 

```{r}
mod <- function() {
  modalDialog(
    tagList(
      textInput(ns("info"), "Your info here")
    ),
    footer = tagList(
      modalButton("Cancel"),
      actionButton(ns("ok"), "OK")
    )
  )
}

observeEvent(input$show, {
  showModal(mod())
})
observeEvent(input$ok, {
  removeModal()
})
```

+ `$('#id').css('color', 'green');` changes the css attributes of the selected element(s). 
Here, we're switching to green on the `#id` element.

+ `$("#id").text( "this" );` changes the text content to "this". 
This can be used to replace 

```{r eval = FALSE}
output$ui <- renderUI({
  if (this){
    tags$p("First")
  } else {
    tags$p("Second")
  }
})
```

+ `$("#id").remove();` completely removes the element from the DOM. 
IT can be used as a replacement for `shiny::removeUI()`, or as a conditional UI. 

### Where to put them - introduction to JavaScript events

OK, now that we've got some ideas about JS code that can be used in Shiny, where do we put them? 
HTML and JS have a concept called `events`, which are... well events that happen when the user manipulates the webpage: when the user clicks, hovers (the mouse goes over an element), press the keyboard... 
All these events can be used to trigger a JavaScript function. 

Here are some example of adding JavaScript functions to DOM events:

+`onclick`

The `onclick` attribute can be add straight inside the HTML tag when possible:

```{r eval = FALSE}
tags$button(
  "Show"
  onclick = "$('#plot').show()"
)
```

Or with `shiny::tagAppendAttributes`:

```{r eval = FALSE}
plotOutput(
  "plot"
) %>% tagAppendAttributes(
  onclick = "alert('hello world')"
)
```

Here is for example a small Shiny app that implements this behavior:

```{r eval = FALSE}
library(shiny)
library(magrittr)
ui <- function(request){
  fluidPage(
    plotOutput(
      "plot"
    ) %>% tagAppendAttributes(
      onclick = "alert('iris plot!')"
    )
  )
}

server <- function(input, output, session){
  output$plot <- renderPlot({
    plot(iris)
  })
}

shinyApp(ui, server)
```

You can find a real Life example of this `tagAppendAttributes` in the `{tidytuesday201942}` app: 

+ [R/mod_dataviz.R#L109](https://github.com/ColinFay/tidytuesday201942/blob/master/R/mod_dataviz.R#L109), where the click on the plot generates the creation of a Shiny input (we'll see this below)

That, of course, works well with very small JavaScript code. 
For longer JavaScript code, you can write a function inside and external file, and add it to your app. 
In `{golem}`, this works by launching the `add_js_file("name")`, which will create a `.js` file. 
You'll then need to add this function using the `tags$script(src="www/name.js")` inside the `golem_add_external_resources()` function in `R/app_ui.R` file. 

This, for example, could be: 

> inst/app/www/script.js 

```{js eval = FALSE, echo = TRUE}
function alertme(id){
  // Asking information
  var name = prompt("Who are you?");
  // Showing an alert
  alert("Hello " + name + "! You're seeing " + id);
}
```

Then:

```{r eval = FALSE}
plotOutput(
  "plot"
) %>% tagAppendAttributes(
  onclick = "alertme('plot')"
)
```

Inside this `inst/app/www/script.js`, you can also attach a new behavior with jQuery to one or several elements. 
For example, you can add this `alertme` / `onclick` behavior to all plots of the app:

```{js eval = FALSE, echo = TRUE}
function alertme(id){
  var name = prompt("Who are you?");
  alert("Hello " + name + "! You're seeing " + id);
}

/* We're adding this so that the function is launched only
when the document is ready */
$(function(){ 
  // Selecting all Shiny plots
  $(".shiny-plot-output").on("click", function(){
    /* Calling the alertme function with the id 
    of the clicked plot */
    alertme(this.id);
  });
});
```

Then, all the plots from your app will receive this on click event^[
This `click` behavior can also be done through `$(".shiny-plot-output").click(...)`. 
We chose to display the `on("click")` pattern as it can be generalized to all DOM events.
].

Note that there is a series of `Shiny events`, which are specific to Shiny but that you can used as the one we've just seen: 

```{js eval = FALSE, echo = TRUE}
function alertme(){
  var name = prompt("Who are you?");
  alert("Hello " + name + "! Welcome to my app");
}

$(function(){ 
  // Waiting for Shiny to be connected
  $(document).on('shiny:connected', function(event) {
    alertme();
  });
});
```

See [JavaScript Events in Shiny](https://shiny.rstudio.com/articles/js-events.html) for the full list of JavaScript events available in Shiny.

## JavaScript <-> Shiny communication 

Now that we've seen some client-side optimization, i.e. R doesn't do anything with these events when they happen (in fact R is not even aware that they happen), let's now see how we can make these two communicate with each other. 

### From R to JavaScript 

Calling JS from the server side (i.e from R) is done by defining a series of `CustomMessageHandler`: these are functions with one argument that can then be called using the `session$sendCustomMessage()` method from the server. 

You can define them using this skeleton:

```{r echo = FALSE, eval = TRUE}
readLines("golex/inst/app/www/plop.js") %>%
  glue::as_glue()
```

This skeleton is the one generated by `golem::add_js_handler("plop")`.

Then, it will be called from server-side with:

```{r eval = FALSE}
session$sendCustomMessage("fun", list())
```

Note that the `list()` argument from your function will be converted to JSON, and read as such from JS. 
In other words, if your have an argument called `x`, and you call the function with `list(a = 1, b = 12)`, then in JS you'll be able to use `x.a` and `x.b`.

For example:

> In `inst/app/www/script.js`

```{js, eval = FALSE, echo = TRUE}
Shiny.addCustomMessageHandler('computed', function(mess) {
  alert("Computed " + mess.what + " in " + mess.sec + " secs");
})
```

Then in R: 

```{r eval = FALSE}
observe({
  deb <- Sys.time()
  # Do the computation for id
  Sys.sleep(
    sample(1:5, 1)
  )
  session$sendCustomMessage(
    "computed", 
    list(
      what = "plop", 
      sec = round(Sys.time() - deb)
    )
  )
})
```

### From JavaScript to R

How to do the other way around (from JS to R)?
Shiny apps, in the browser, contain an object called `Shiny`, which can be used to send values to R, by creating an `IputValue`. 
For example, with: 

```{js, eval = FALSE, echo = TRUE}
Shiny.setInputValue("rand", Math.random())
```

you'll bind an input that can be caught from the server side with:

```{r eval = FALSE}
observeEvent( input$rand , {
  print( input$rand )
})
```

This `Shiny.setInputValue` can of course be used inside any JavaScript function. 
Here is a small example wrapping some of the things we've seen previously:

> In `inst/app/www/script.js`

```{js eval = FALSE, echo = TRUE}
function alertme(){
  var name = prompt("Who are you?");
  alert("Hello " + name + "! Welcome to my app");
  Shiny.setInputValue("username", name)
}

$(function(){ 
  // Waiting for Shiny to be connected
  $(document).on('shiny:connected', function(event) {
    alertme();
  });
  
  $(".shiny-plot-output").on("click", function(){
    /* Calling the alertme function with the id 
    of the clicked plot */
    Shiny.setInputValue("last_plot_clicked", this.id);
  });
});
```

These events (getting the user name and the last plot clicked), can then be caught from the server side with:

```{r eval = FALSE}
observeEvent( input$username , {
  cli::cat_rule("User name:")
  print(input$username)
})

observeEvent( input$last_plot_clicked , {
  cli::cat_rule("Last plot clicked:")
  print(input$last_plot_clicked)
})
```

Which will give:

``` 
> golex::run_app()
Loading required package: shiny

Listening on http://127.0.0.1:5495
── User name: ─────────────────────────────────────────────────────
[1] "Colin"
── Last plot clicked: ─────────────────────────────────────────────
[1] "plota"
── Last plot clicked: ─────────────────────────────────────────────
[1] "plopb"
```
__Important note__: if you're using modules, you'll need to pass the namespacing of the `id` to be able to get it back from the server. 
This can be done using the `session$ns` function, which comes by default in any golem-generated module. 
And as an example speaks more than a 1000 words, here is what we mean by that:

```{js eval = FALSE, echo = TRUE}
$( document ).ready(function() {
  Shiny.addCustomMessageHandler('whoareyou', function(arg) {
    var name = prompt("Who are you?")
    Shiny.setInputValue(arg.id, name);
  })
});
```

```{r}
mod_my_first_module_ui <- function(id){
  ns <- NS(id)
  tagList(
    actionButton(
      ns("showname"), "Enter your name"
    )
  )
}

mod_my_first_module_server <- function(input, output, session){
  ns <- session$ns
  observeEvent( input$showname , {
    session$sendCustomMessage(
      "whoareyou", 
      list(
        id = ns("name")
      )
    )
  })
  
  observeEvent( input$name , {
    cli::cat_rule("Username is:")
    print(input$name)
  })
}
```


## About `{golem}` js functions

`{golem}` comes with a series of JavaScript functions that you can call from the server. 
These functions are added by default with `golem::activate_js()` in `app_ui`.

Then they are called with `golem::invoke_js("function", "element")`.

This `element` is a series of elements (most of the time scalar elements) that can be used to select the DOM node you need to interact with.
It can be a full jQuery selector, an id or a class.
Note that you can pass multiple elements, with `invoke_js` `...` parameters

### `golem::invoke_js()`

+ `showid` & `hideid`, `showclass` & `hideclass` show and hide elements through their id or class.

```{r eval = FALSE}
golem::invoke_js("showid", ns("plot"))
```

+ `showhref` & `hidehref` hide and show a link by trying to match the `href` content.

```{r eval = FALSE}
golem::invoke_js("showhref", "panel2")
```

+ `clickon` click on the element. You have to use the full jQuery selector. 

+ `show` & `hide` show and hide elements, using the full jQuery selector.

See `?golem::activate_js` for a full list of built-in functions.

## Learn more about JavaScript

Here is a list of external resources to learn more about JavaScript: 

### JavaScript basics 

+ [Mozilla JavaScript](https://developer.mozilla.org/en-US/docs/Web/JavaScript)
+ [w3schools JavaScript](https://www.w3schools.com/js/default.asp)
+ [Free Code Camp](https://www.freecodecamp.org/)
+ [JavaScript For Cats](http://jsforcats.com/)
+ [Learn JS](https://www.learn-js.org/)

### jQuery

+ [jQuery Learning Center](https://learn.jquery.com/)
+ [w3schools jQuery](https://www.w3schools.com/jquery/default.asp)

### Intermediate / advanced JavaScript

+ [Eloquent JavaScript](https://eloquentjavascript.net/)
+ [You Don't Know JS Yet](https://github.com/getify/You-Dont-Know-JS)
