---
title: "Use renv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE
)
```

# First time

- Be sure to be inside your Rstudio project

## You may need {renv}

```{r}
remotes::install_github('rstudio/renv')
```

## Hide from git
```{r}
# Should already be done
# install.packages("usethis")
usethis::use_git_ignore("renv_instructions.Rmd")
usethis::use_build_ignore("renv_instructions.Rmd")
```

## The first time you launch your project in the Docker container, run :
```{r}
renv::consent(TRUE)
renv::activate()
```

## Restore the project from the lock file if any problem
```{r}
renv::init()
```



# Activate cache shared on host
- Store this code in your *project* `.Rprofile` file
  `Sys.setenv(RENV_PATHS_CACHE = "/opt/local/renv/cache")`
- If cache *not* shared with host with option `renv_cache`, then de-activate cache otherwise packages will disappear at next Docker restart
`renv::settings$use.cache(FALSE)`

## .Rprofile content
Store this code inside your project `.Rprofile` file to reduce problems with cache.

```r
source("renv/activate.R")
renv::activate()

# cache ----
if (dir.exists("/opt/local/renv/cache")) {
  # Cache inside the docker container with persistent drive
  Sys.setenv(RENV_PATHS_CACHE = "/opt/local/renv/cache")
  renv::settings$use.cache(TRUE)
} else if (dir.exists("/mnt/Data/renv_cache")) {
  # Your local path on linux/unix if project used out of Docker
  Sys.setenv(RENV_PATHS_CACHE = "/mnt/Data/renv_cache")
  renv::settings$use.cache(TRUE)
} else if (dir.exists("D:/renv_cache_windows/")) {
  # Your local path on Windows if project used out of Docker
  Sys.setenv(RENV_PATHS_CACHE = "D:/renv_cache_windows")
  renv::settings$use.cache(TRUE)
} else {
  # No cache
  renv::settings$use.cache(FALSE)
}
```

# During the dev process
## After you updated packages, added new ones or removed somes
If you are satisfied with the effects, run:
```{r}
renv::snapshot()
```

If you are not satisfied, run:
```{r}
renv::restore() # instead of snapshot()
```

## If you are ready to send your modifications to the git server
And after you ran `devtools::check()`, store your packages list with:

```{r}
renv::snapshot()
```

## If you updated your branch from the server
There may be new packages needed, therefore run:
```{r}
renv::restore()
```

# If you need to upgrade (or otherwise change) the version of renv
This should normally be automatically done in the `renv/activate.R`, but you can do it with:

```{r}
renv::upgrade()
```


# Use git inside the Docker container
## First time
```{r}
# project config
usethis::use_git_config(scope = "project", user.name = "username", user.email = "user@email.fr")
```

## Use credentials
Must be done for each session as not stored in the persistent drive

```{r}
# Store credentials
my_cred <- git2r::cred_user_pass(
  username = "username",
  password = askpass::askpass()
)
usethis::use_git_credentials(credentials = my_cred)
```

# If you mistakenly installed package from github or forced one version
You may have to be sure to install dependencies from MRAN
```{r}
ll <- list.files(.libPaths()[1], full.names = TRUE)
for (l in ll) {
  try(remotes::install_cran(l))
}
```

